Space is the Place: Effects of Continuous Spatial Structure on Analysis of Population Genetic Data

### 2019-11-22 ms/review-responses.tex

\begin{point}{Figure S3:}
    Caption seems to be missing detail 
\end{point}

\reply{
    Thanks for catching this - we have revised this caption to add details including the simulation parameters. (Supplemental Figure \ref{fig:hetmap}).
}

%%%%%%%%%%%%%%
\reviewersection{1}

\begin{quote}
    % This study explores biases arising in population-based inference when 1)
    % real population samples are coming from spatial habitat with various degree
    % of structuring while inference is made assuming random mating population;
    % 2) imperfect sampling in practice that fails to represent full diversity
    % across entire population habitat; 3) phenotypes that vary across geography
    % and create spurious associations with genotypes. While earlier studies
    % explored the effect of strong structure on population genetic inference and
    % GWAS, this work focuses on less extreme scenarios of structuring that
    % arises in populations evolving in continuous habitat. By using
    % non-Wright-Fisher model, authors simulated chromosome-scale samples from
    % populations that evolved in continuous space, and that can model
    % environmental factors to create phenotypes varying over space. As a result,
    % this study identified spatial structuring scenarios (small neighbourhood
    % size $\sim$ 10--100) that coupled with imperfect sampling strategies lead to a
    % biased inference of widely used population genetic statistics (altogether
    % 18 statistics) such as pi (average pairwise sequence differences),
    % heterozygosity (and inbreeding coefficient), and IBS tract sharing.
    % Accordingly, inference of the effective population size history was also
    % strongly affected under these parameter ranges. Finally, the authors use
    % their spatial modelling to demonstrate that typical GWAS with PC-based
    % correction cannot entirely remove spurious signals of genotype-phenotype
    % associations arising from purely environmental factors.
    Overall, the
    authors explore an important but often neglected source of bias that can
    affect inference in many population-based studies (in medical genetics,
    evolutionary biology and ecology). This study can be of interest to a
    broader audience of readership, and I have only minor comments to improve
    clarity and increase accessibility for readers:
\end{quote}

@@

\begin{point}{}
    When neighbourhood size is small (10-100), the mean number of IBS tracts ${}>2$bp (nIBS as in Table S1) is elevated similar to Wright's inbreeding coefficient, but mean of the distribution of pairwise IBS (mean(IBS)) is decreased. What could be the source of this discrepancy? How exactly mean(IBS) was calculated?
\end{point}

\reply{
    The mean IBS is simply the mean of the vector of lengths of IBS tracts.
    \plr{make sure this is stated in the text somewhere}
    The short answer is that if one splits a chromosome up into more pieces (increasing nIBS),
    the mean length of those pieces must necessarily be smaller.
    At low neighborhood sizes, spatially distant individuals tend to have longer coalescence times,
    leading to more (and hence smaller) IBS tracts.
    This can be seen in Supplemental Figure \ref{fig:allsumstats}.
    Note that skew(IBS) and number of long ($>$1e6) IBS tracts reflect other aspects of the distribution.
}

\begin{point}{}
    The authors use K to denote both carrying capacity \llname{ll:K_defn}
    and population density \llname{ll:K_again}. 
    It might be better to use a different notation for these quantities since carrying capacity is fixed while density is an emergent quantity in the non-Wright-Fisher model. Use of K to denote carrying capacity and density is a bit confusing. For example, on \llname{ll:K_more} it is said that ' the ``population density'' (K) and ``mean lifetime'' (L) parameters were the same in all simulations'. Here K seems to indicate carrying capacity rather than density? The latter is an emergent quantity and varies across simulation runs?
\end{point}

\reply{
    We agree and have
    adjusted our language to underscore the point that
    $K$ is a parameter that controls population density, rather than being equal to it,
    at \llname{ll:K_again} and \revref{} and \llname{ll:K_more}.
}

@@

\begin{point}{}
    To what extent are any of the results dependent on the exact method of simulation. There are a number of choices about the exact details of the simulations (e.g. the way the overlapping generations are handled, the edge effects and, particularly, the form of Equation 1 - see below). It's not so much that these are non-standard (since I don't think there is a standard) and they all sort of make sense heuristically, and I was left wondering whether these sorts of choices actually make a difference. Do the authors have some thoughts/intuition/results about that? Given that the results in \autoref{fig:sumstats} seem quite consistent with expectations, I suspect that on some level it doesn't make much difference but then there are intermediate results like \autoref{fig:genparams} which seem a bit counter-intuitive and I wonder if those aspects depend on the simulation scheme.
\end{point}

\reply{
    This is a good point also raised by other reviewers.
    We have added some more discussion of the choice of demographic model
    (\revreffull{AE}{2} and \llname{demog_disco}).
}

\begin{point}{}
    Related to the first point, to what extent are the results qualitatively different to those that would be obtained in a stepping-stone model? My interpretation is that they are actually very similar, but I didn't see whether that was explicitly discussed. In some sense, it's still easier to do large simulations in a stepping-stone model so it would be nice to be reassured that that's still ok.
\end{point}

\reply{
    We have added an appendix comparing our model to a reverse-time stepping stone simulation on several relevant summary statistics \llname{apx:stepping_stone}.
    For some of these statistics (e.g., $\theta_\pi$),
    the behavior of the stepping stone model becomes more similar (although not identical)
    to the continuous model as the resolution of the landscape increases,
    but other statistics are more strongly affected by discretization.
    We are also curious how all the other analyses we perform would be affected using other simulation schemes, but hope to explore that aspect in future work. 
    Furthermore, discrete-deme coalescent simulations are certainly faster,
    but come with significant issues either when neighborhood size is lower than the population of a deme,
    or when demes are small enough that the sample size approaches local $N_e$.
}

@@

We are glad you agree this is an important topic to explore!

\begin{quote}
    Although I was enthusiastic about the topic, I was quite disappointed with
    the core of the study, i.e., the forward-in-time simulation of populations in
    continuous space. The field has been struggling with this issue for decades --
    examples of spectacular failures like the Wright-Malecot model (see
    Felsenstein's ``pain in the torus'' article, 1975) or, more recently, the
    ``mugration'' or ``discrete trait analysis'' model in phylodynamics (see De Maio
    et al. 2015) have probably mostly harmed our research field -- that one cannot
    make the economy of using a sound probabilistic model for generating
    geo-referenced genetic data. It does not seem to be the case here
    unfortunately.  
\end{quote}

We were surprised that the reviewer seems to be saying that local, density-mediated control of mortality
doesn't constitute a ``sound probabilisitic model'',
but hope that our additional discussion of the choice,
and citations to the history of this sort of model,
help to better motivate our choice.

\begin{point}{}
    First, the simulation starts with individuals distributed uniformly at random in space. Is there any indication that the three-step algorithm used here maintains this distribution during the course of evolution? If it does not, then is there any stationary regime and how many generations does one need to wait before reaching it? I do appreciate that the competitive interaction term was introduced in order to avoid seeing the ``clumping'' of individuals that hampers the Wright-Malecot model. Yet, just because there are no such clusters does not mean that the spatial distribution of individuals reaches a stable regime and that the distribution reached, if any, is reasonable from a biological perspective.
\end{point}

@@

\begin{point}{}
    All in all, more efforts need to be made here in my opinion to show that the forward-in-time simulations generate sensible outcomes. Sensible in terms of the behavior of the population demography at equilibrium (provided such equilibrium indeed exists) along with that of the spatial distribution of individuals. The authors could provide some guarantee of the good behavior of their model as evidenced from simulations using a broad range of parameter values for generating data. Alternatively, they could elect to use the spatial-Lambda-Fleming-Viot model for their simulations, which, in my opinion would seem the most sensible option given that (1) it is possible to run backward-in-time simulations under this model, thereby saving a lot of computation time and (2) it is a well-studied model with good mathematical and biological properties and (3) it is implemented in a publicly available software program (\url{https://github.com/jeromekelleher/discsim}).
\end{point}

\reply{
    Hopefully, the additional plots 
    and discussion around demographic modeling (notably, the discussion at \llname{demog_disco})
    helps to increase the transparency in this model.
    We agree that demographic realism is a very important consideration,
    and due to the often hard-to-intuit nature of spatial demography,
    simulations need to be carefully sanity checked.
    (Furthermore, what is ``realistic'' for one species will not be for another!)
    We have added an Appendix comparing our model to a stepping-stone \llname{apx:stepping_stone}, 
    because we think these are the most familiar and widely used class of spatial models.
    We find that many features of our model are well approximated by stepping-stone models, and that for statistics like $\theta_{\pi}$ the stepping stone model results approach our continuous space model as the number of demes used to describe the landscape increases. 
}

    We definately think that a comparison to the spatial Lambda-Fleming-Viot
    would be an extremely useful thing to do, for the reasons the reviewer mentions,
    but this is beyond the scope of the current study.
    The SLFV model is only known to be an approximation of one particular biologically explicit model
    (of patchy extinction-recolonization)
    and is only conjectured to be a good approximation
    for spatial models more generally.
    To reiterate, the definition of the SLFV was explicitly motivated by a desire to approximate
    spatial models, especially Bolker-Pacala models like ours.
    Therefore, our study is an important prerequisite
    for simulation study of the spatial Lambda-Fleming-Viot,
    as it works through some of the issues in simulating biologically realistic spatial models,
    and describes some of the patterns.

@@

\begin{point}{}
    Demographic modeling. Both approaches tested, stairwayplot and SMC++, are most sensitive to older demographic events, and consequently are very noisy and underestimate effect population sizes, especially in smaller neighborhood sizes. Models that consider haplotype structure are much better suited to this time period. It would be helpful to either 1) discuss the varying time sensitivities of different classes of demographic inference methods and how spatial patterns of genetic variation would influence these inferences, or 2) apply a method of this class (many options, e.g. DoRIS, IBDNe, Tracts, Globetrotter, etc) and show how it performs.
\end{point}

\reply{
    We now discuss haplotype methods in the relevant discussion section \llname{ibdnesxn}. However though these methods should be more accurate for recent events it is not clear that this will improve performance per se. The dips in recent inferred Ne from stairwayplot are not just prediction noise, but actually reflect an underlying genealogy in which terminal branches are shorter than expected from a constant-size random-mating population (see e.g. figure \ref{fig:sumstats}A and \ref{fig:ibs}). The interpretation error is that these short branches are generated by spatial structure rather than changes in population size over time -- a point also made in the \citep{Mazet2015} paper we discuss in the introduction and discussion. 
}

\begin{point}{}
    GWAS mixed models. To what extent can spatial signals (e.g. corner, patchy) be corrected with mixed models, e.g. with PCs and PC-adjusted GRM as in Conomos et al, 2016 using PC-AiR and PC-Relate)? Is patchiness related to dispersal? I'm curious how this relates to the predictive ability of GWAS phenotypes with some spatial association that may or may not be associated with environmental effects.
\end{point}

\reply{
    Good question -- we are also interested to know how mixed models perform here, but think that to properly test that we would want to change our design to generate phenotypes from simulated genotypes. This would allow us to evaluate false-negatives in addition to false-positives. This is important because, if mixed models do provide stronger control for stratification they are also likely to remove true signals of causal SNPs if those SNPs covary with spatial structure. We now point to these methods explicitly in the discussion \llname{mixedmodelcallout}, but think that incorporating that study here would make this paper too long and we plan to explore this timely issue in the future. We also think the PC results are still quite relevant as the method is quite population in current studies. 
}

@@

\begin{point}{}
    Definitions and interpretations. There are quite a large number of metrics discussed in Figure 3B, and it's a lot to take in. It might be helpful to have a table with a reminder of what the metric is, its interpretation, and how it is computed.
\end{point}

\reply{
    We have included a table describing the summary statistics in Figure \ref{table:sumstats}.
}

\begin{point}{}
    Notation: ``Offspring disperse a Gaussian-distributed distance away from the parent with mean zero and standard deviation $\sigma$ in both the x and y coordinates. Each offspring is produced with a mate selected randomly from those within distance $3\sigma$, with probability of choosing a neighbor at distance $x$ proportional to $\exp(-x^2/2\sigma^2)$.'' I think x may be overloaded here, or I'm confused. Clarify?
\end{point}

\reply{
	We have clarified the notation for the second instance (referring to the distance among individuals) to $d$. 
}

### 2019-11-22 ms/spaceness.tex
took a pass on the main text

To avoid hard-to-evaluate approximations, we here used forward-time, individual-based simulations.
The question of what regulates real populations has a long history and many answers
\citep[e.g.,][]{lloyd1967crowding,antonovics1980ecological,crawley1990population}, \llabel{ecolcites}
but it is clear that populations must at some point have density-dependent feedback on population size,
or else they would face eventual extinction or explosion.
In the absence of unrealistic global population regulation,
this regulation must be local,
and there are many ways to arrange this \citep{bolker2003spatial}.
In our simulations,
we decrease each individual's probability of survival with increasing local population density,
which shifts reproductive output towards low-density regions,
and produces total census sizes that fluctuate around an equilibrium. \revpoint{3}{2}
This also prevents the population clumping seen by \citet{Felsenstein1975}
(Supplemental \autoref{fig:wm_v_dd})). \revpoint{3}{1}
Such models have been used extensively in ecological modeling
\citep{durrett1994importance,bolker1997using,law2003population,fournier2004microscopic,champer2019suppression}
but rarely in population genetics, \revpoint{AE}{2}
where to our knowledge implementations of continuous space models
before their availability through SLiM \citep{Haller2019}
have focused on a small number of genetic loci
\citep[e.g.,][]{slatkin1989comparison,barton2002neutral,robledoarnuncio2010isolation,rossine2014espao},
which limits the ability to investigate the impacts of continuous space on genome-wide genetic variation as is now routinely sampled from real organisms. 
By simulating chromosome-scale sequence alignments and complete population histories we are able to treat our simulations as real populations and replicate the sampling designs and analyses commonly conducted on real genomic data.

@@

We chose our demographic model so that every individual has on average $1/L$ offspring
each time step,
and if the local population density of an individual is $n$,
then their probability of survival until the next time step is
(equation \eqref{eqn:p_survival}):
\begin{equation}
    p = \min\left( 0.95, \frac{1}{1 + n / (K (1+L))} \right) .
\end{equation}
We capped survival at 0.95 so that we would not have exceptionally long-lived individuals
in sparsely populated areas --
otherwise, an isolated individual might live for a very long time.
Since $1 - p \approx n / (K (1 + L))$,
mortalitity goes up roughly linearly with the number of neighbors (on a scale given by $K$),
as would be obtained if, for instance, mortality is due to agonistic interactions.
Ignoring in/outmigration,
a region is at demographic equilibrium if the per-capita probability of death
is equal to the birth rate,
i.e., if $1 - p = 1/L$.
(Note that there is no effect of age in the model,
which would make the analysis more complicated.)
Solving this for $n$,
we get that in a well-mixed population, the equilibrium density should be around
\begin{equation}
    n = K \frac{L+1}{L-1}
\end{equation}
individuals per unit area.
At this density, the per-capita death rate is $1/L$, so the mean lifetime is $L$.
This equilibrium density is \emph{not} $K$, but (since $L=4$) is two-thirds larger.
However, in practice this model leads to a total population size which is around $K$
multiplied by total geographic area
(but which depends on $\sigma$, as discussed above).
The main reason for this is that
since offspring tend to be near their parents, individuals tend to be ``clumped'',
and so experience a higher average density than the ``density'' one would compute
by dividing census size by geographic area \citep{lloyd1967crowding}.
To maintain a constant expected total population size
would require making (say) $K$ depend on $\sigma$;
however, typical local population densities might then be more dissimilar.

>>>>

To avoid hard-to-evaluate approximations, we here used forward-time, individual-based simulations.
The question of what regulates real populations has a long history and many answers
\citep[e.g.,][]{lloyd1967crowding,antonovics1980ecological,crawley1990population}, \llabel{ecolcites}
but it is clear that populations must at some point have density-dependent feedback on population size,
or else they would face eventual extinction or explosion.
In the absence of unrealistic global population regulation,
this regulation must be local,
and there are many ways to achieve this \citep{bolker2003spatial}.
In our simulations,
we decrease each individual's probability of survival with increasing local population density,
which shifts reproductive output towards low-density regions,
and produces total census sizes that fluctuate around an equilibrium. \revpoint{3}{2}
This also prevents the population clumping seen by \citet{Felsenstein1975}
(Supplemental \autoref{fig:wm_v_dd})). \revpoint{3}{1}
Such models have been used extensively in ecological modeling
\citep{durrett1994importance,bolker1997using,law2003population,fournier2004microscopic,champer2019suppression}
but rarely in population genetics, \revpoint{AE}{2}
where to our knowledge implementations of continuous space models
before their availability through SLiM \citep{Haller2019}
have focused on a small number of genetic loci
\citep[e.g.,][]{slatkin1989comparison,barton2002neutral,robledoarnuncio2010isolation,rossine2014espao},
which limits the ability to investigate the impacts of continuous space on genome-wide genetic variation as is now routinely sampled from real organisms. 
By simulating chromosome-scale sequence alignments and complete population histories we are able to treat our simulations as real populations and replicate the sampling designs and analyses commonly conducted on real genomic data.

@@

We chose our demographic model so that every individual has on average $1/L$ offspring
each time step,
and if the local population density of an individual is $n$,
then their probability of survival until the next time step is
(equation \eqref{eqn:p_survival}):
\begin{equation}
    p = \min\left( 0.95, \frac{1}{1 + n / (K (1+L))} \right) .
\end{equation}
We capped survival at 0.95 so that we would not have exceptionally long-lived individuals
in sparsely populated areas --
otherwise, an isolated individual might live for a very long time.
Since $1 - p \approx n / (K (1 + L))$,
mortality goes up roughly linearly with the number of neighbors (on a scale given by $K$),
as would be obtained if, for instance, mortality is due to agonistic interactions.
Ignoring migration,
a region is at demographic equilibrium if the per-capita probability of death
is equal to the birth rate,
i.e., if $1 - p = 1/L$.
(Note that there is no effect of age in the model,
which would make the analysis more complicated.)
Solving this for $n$,
we get that in a well-mixed population, the equilibrium density should be around
\begin{equation}
    n = K \frac{L+1}{L-1}
\end{equation}
individuals per unit area.
At this density, the per-capita death rate is $1/L$, so the mean lifetime is $L$.
This equilibrium density is \emph{not} $K$, but (since $L=4$) is two-thirds larger.
However, in practice this model leads to a total population size which is around $K$
multiplied by total geographic area
(but which depends on $\sigma$, as discussed above).
The main reason for this is that
since offspring tend to be near their parents, individuals tend to be ``clumped'',
and so experience a higher average density than the ``density'' one would compute
by dividing census size by geographic area \citep{lloyd1967crowding}.
To maintain a constant expected total population size
would require making (say) $K$ depend on $\sigma$;
however, typical local population densities might then be more dissimilar.
